---
title: "visualization of TF activity"
author: "Cong"
date: "2024-07-01"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# import packages and functions
```{r import packages and functions}
suppressMessages(library(data.table))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(xlsx))
suppressMessages(library(pheatmap))
suppressMessages(library(RColorBrewer))


set.seed(42)
fl.sources <- list.files("../../scripts/utils/", full.names = T)
tmp <- sapply(fl.sources,source)
```

# set up plot parameters
```{r plot set-up}
## annotation file
group_sorted <- read.csv('group_file_for_heatmap.csv',header = T, row.names = 1)
mycolors <- list(subsets = c("#A6A6A6", "#FF0000", "#0007F4", "#FFA75C", "#00AAFE", 
                             "#009051", "#FEB5B5", "#E6CFFF", "#8AAA75"))
names(mycolors$subsets) <- unique(group_sorted$Group)
breaksList = seq(-2, 2, by = 0.1)


# function
get_heatmap <- function(df,size=3,...){
  g2 <- group_sorted[colnames(df),]
  annotation_col = data.frame(subsets = g2$Group) 
  rownames(annotation_col) = colnames(df)
  p1 <- pheatmap(df, 
                 fontsize = size, angle_col = 90, 
                 cellwidth = 2*size, cellheight = size,
                 cluster_cols = F, 
                 annotation_col = annotation_col,
                 clustering_distance_cols = 'correlation', 
                 clustering_distance_rows = 'correlation', 
                 clustering_method = 'average',
                 # annotation_row = annotation_row,
                 breaks = breaksList,
                 color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)),
                 annotation_names_row = F, annotation_names_col = F,
                 annotation_colors = mycolors, border_color = NA, ...)
  return(p1)
}

```


# heatmap visualization for multi-tasker and single-tasker
Each row represents TF and each column represents sample. The upper color annotation represents the cell type of each sample. Each cell is the PageRank score with red displaying high scores.

```{r import data}
## import data
df_multi <- read.xlsx("Extended Data Table 2 _ TF PageRank Scores of Multi versus Single state specific TFs.xlsx", sheetIndex = 1, row.names=T) 
df_single <- read.xlsx("Extended Data Table 2 _ TF PageRank Scores of Multi versus Single state specific TFs.xlsx", sheetIndex = 2, row.names=T) 
```


TFs active in multi-states
```{r get multi-tasker heatmap, fig.width=5, fig.height=8}
## get heatmap
p1 <- get_heatmap(df_multi)
```

TFs activity enriched in a single state
```{r get single-tasker heatmap, fig.width=5, fig.height=8}
p2 <- get_heatmap(df_single, cluster_rows=F)
```

# heatmap of non-specific TFs
```{r}
df <- read.xlsx("Extended Data Table 1 _ TF PageRank Scores of non-specific versus specific TFs.xlsx", sheetIndex = 1, row.names=T)
get_heatmap(df, cluster_rows=F) 
```


# heatmap of cell state-specific TFs
From top to bottom, they're Naive, MP, TE, TCM, TEM, TRM, TexProg, TexInt, and TexTerm.
```{r}
dfs <- lapply(2:10, function(x) read.xlsx("Extended Data Table 1 _ TF PageRank Scores of non-specific versus specific TFs.xlsx", sheetIndex = x, row.names=T))

ps <- lapply(1:9, function(x) {
  annotation_row <- data.frame("subsets"=rep(names(mycolors$subsets)[x],nrow(dfs[[x]])))
  rownames(annotation_row) <- rownames(dfs[[x]])
  get_heatmap(df=dfs[[x]], cluster_rows=F, annotation_row=annotation_row) 
})
```



# bubble visualization for TRM and TEXterm TFs
```{r}
## import data
dfs <- lapply(1:3, function(x) {
  o <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = x, row.names=T, startRow = 2)
  o <- o[,grep('TRM|TexTerm',names(o))]
  return(list(o[,1:9], o[,10:18]))
})

# get heatmap
gs <- lapply(dfs, function(x) bubble(rownames(x[[1]]), x[[2]], x[[1]]))

```


TFs that most active in TEXterm
```{r, fig.width=4, fig.height=8}
plot(gs[[1]])
```

TFs that most active in TRM
```{r, fig.width=4, fig.height=5}
plot(gs[[2]])
```

TFs that  active in TRM and TEXterm
```{r, fig.width=4, fig.height=7}
plot(gs[[3]])
```



# session info
```{r }
sessionInfo()
```

