---
title: "visualization of TF activity"
author: "Cong"
date: "2024-07-01"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# import packages and functions
```{r import packages and functions}
suppressMessages(library(data.table))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(xlsx))
suppressMessages(library(pheatmap))
suppressMessages(library(RColorBrewer))


set.seed(42)
fl.sources <- list.files("../../scripts/utils/", full.names = T)
tmp <- sapply(fl.sources,source)
```

# set up plot parameters
```{r plot set-up}
## annotation file
group_sorted <- read.csv('group_file_for_heatmap.csv',header = T, row.names = 1)
mycolors <- list(subsets = c("#A6A6A6", "#FF0000", "#0007F4", "#FFA75C", "#00AAFE", 
                             "#009051", "#FEB5B5", "#E6CFFF", "#8AAA75"))
names(mycolors$subsets) <- unique(group_sorted$Group)
breaksList = seq(-2, 2, by = 0.1)

# function
get_heatmap <- function(df,size=3,...){
  g2 <- group_sorted[colnames(df),]
  annotation_col = data.frame(subsets = g2$Group) 
  rownames(annotation_col) = colnames(df)
  p1 <- pheatmap(df, 
                 fontsize = size, angle_col = 90, 
                 cellwidth = 2*size, cellheight = size,
                 cluster_cols = F, 
                 annotation_col = annotation_col,
                 clustering_distance_cols = 'correlation', 
                 clustering_distance_rows = 'correlation', 
                 clustering_method = 'average',
                 # annotation_row = annotation_row,
                 breaks = breaksList,
                 color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)),
                 annotation_names_row = F, annotation_names_col = F,
                 annotation_colors = mycolors, border_color = NA, ...)
  return(p1)
}

```


# heatmap visualization for multi-tasker and single-tasker
Each row represents TF and each column represents sample. The upper color annotation represents the cell type of each sample. Each cell is the PageRank score with red displaying high scores.

```{r import data}
## import data
df_multi <- read.xlsx("Extended Data Table 2 _ TF PageRank Scores of Multi versus Single state specific TFs.xlsx", sheetIndex = 1, row.names=T) 
df_single <- read.xlsx("Extended Data Table 2 _ TF PageRank Scores of Multi versus Single state specific TFs.xlsx", sheetIndex = 2, row.names=T) 
```


```{r get multi-tasker heatmap, fig.width=5, fig.height=8}
## get heatmap
p1 <- get_heatmap(df_multi)
print(p1)
```


```{r get single-tasker heatmap, fig.width=5, fig.height=8}
p2 <- get_heatmap(df_single, cluster_rows=F)
print(p2)    
```

```{r save to file}
## save to file
pdf("hp_multitasker.pdf")
print(p1)
dev.off()

pdf("hp_singletasker.pdf")
print(p2)
dev.off()

```
# bubble visualization for TRM and TEXterm TFs
```{r}
## import data
df_tex <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = 1, row.names=T, startRow = 2)
df_tex <- df_tex[,grep('TRM|TexTerm',names(df_tex))]
df_tex_pr <- df_tex[,1:9]
df_tex_rna <- df_tex[,10:18]

df_trm <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = 2, row.names=T, startRow = 2)
df_trm <- df_trm[,grep('TRM|TexTerm',names(df_trm))]
df_trm_pr <- df_trm[,1:9]
df_trm_rna <- df_trm[,10:18]

df_both <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = 3, row.names=T, startRow = 2) 
df_both <- df_both[,grep('TRM|TexTerm',names(df_both))]
df_both_pr <- df_both[,1:9]
df_both_rna <- df_both[,10:18]

```

```{r}
g1 <- bubble(rownames(df_tex), df_tex_rna, df_tex_pr)
g2 <- bubble(rownames(df_trm), df_trm_rna, df_trm_pr)
g3 <- bubble(rownames(df_both), df_both_rna, df_both_pr)
```

TFs that most active in TEXterm
```{r, fig.width=4, fig.height=8}
plot(g1)
```

TFs that most active in TRM
```{r, fig.width=4, fig.height=5}
plot(g2)
```

TFs that  active in TRM and TEXterm
```{r, fig.width=4, fig.height=7}
plot(g3)
```

```{r}
## save to file
pdf("bb_TEX_active_34TFs.pdf", width=4, height=8)
grid.draw(g1)
dev.off() 

pdf("bb_TRM_active_20TFs.pdf", width=4, height=5)
grid.draw(g2)
dev.off() 

pdf("bb_TEX_and_TRM_30TFs.pdf",width=4, height=7)
grid.draw(g3)
dev.off() 
```



# session info
```{r }
sessionInfo()
```

