---
title: "visualization of TF activity"
author: "Cong"
date: "2024-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## import packages and functions
```{r import packages and functions}
suppressMessages(library(data.table))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(xlsx))
suppressMessages(library(pheatmap))
suppressMessages(library(RColorBrewer))


set.seed(42)
fl.sources <- list.files("../../scripts/utils/", full.names = T)
tmp <- sapply(fl.sources,source)
```

## set up plot parameters
```{r plot set-up}
## annotation file
group_sorted <- read.csv('group_file_for_heatmap.csv',header = T, row.names = 1)
mycolors <- list(subsets = c("#A6A6A6", "#FF0000", "#0007F4", "#FFA75C", "#00AAFE", 
                             "#009051", "#FEB5B5", "#E6CFFF", "#8AAA75"))
names(mycolors$subsets) <- unique(group_sorted$Group)
breaksList = seq(-2, 2, by = 0.1)

# function
get_heatmap <- function(df,...){
  g2 <- group_sorted[colnames(df),]
  annotation_col = data.frame(subsets = g2$Group) 
  rownames(annotation_col) = colnames(df)
  p1 <- pheatmap(df, 
                 fontsize = 3, angle_col = 90, 
                 cellwidth = 4, cellheight = 3,
                 cluster_cols = F, 
                 annotation_col = annotation_col,
                 clustering_distance_cols = 'correlation', 
                 clustering_distance_rows = 'correlation', 
                 clustering_method = 'average',
                 # annotation_row = annotation_row,
                 breaks = breaksList,
                 color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)),
                 annotation_names_row = F, annotation_names_col = F,
                 annotation_colors = mycolors, border_color = NA, ...)
  return(p1)
}

```


## heatmap visualization for multi-tasker and single-tasker
Each row represents TF and each column represents sample. The upper color annotation represents the cell type of each sample. Each cell is the PageRank score with red displaying high scores.

```{r import data}
## import data
df_multi <- read.xlsx("Extended Data Table 2 _ TF PageRank Scores of Multi versus Single state specific TFs.xlsx", sheetIndex = 1, row.names=T) 
df_single <- read.xlsx("Extended Data Table 2 _ TF PageRank Scores of Multi versus Single state specific TFs.xlsx", sheetIndex = 2, row.names=T) 
```


```{r get heatmap}
## get heatmap
p1 <- get_heatmap(df_multi)
print(p1)
```


```{r get heatmap}
p2 <- get_heatmap(df_single, cluster_rows=F)
print(p2)    
```

```{r save to file}
## save to file
pdf("hp_multitasker.pdf")
print(p1)
dev.off()

pdf("hp_singletasker.pdf")
print(p2)
dev.off()

```
## bubble visualization for TRM and TEXterm TFs
```{r}
## import data
df_tex <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = 1, row.names=T, startRow = 2)
df_tex_pr <- df_tex[,1:42]
df_tex_rna <- df_tex[,43:84]

df_trm <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = 2, row.names=T, startRow = 2)
df_trm_pr <- df_trm[,1:42]
df_trm_rna <- df_trm[,43:84]

df_both <- read.xlsx("Extended Data Table 3 _ TF PageRank Scores of TRM and TEXterm TFs.xlsx", sheetIndex = 3, row.names=T, startRow = 2) 
df_both_pr <- df_both[,1:42]
df_both_rna <- df_both[,43:84]

```

```{r}
g1 <- bubble(rownames(df_tex), df_tex_rna, df_tex_pr)
g2 <- bubble(rownames(df_trm), df_trm_rna, df_trm_pr)
g3 <- bubble(rownames(df_both), df_both_rna, df_both_pr)
```


```{r}
## save to file
pdf("bb_TEX_active_34TFs.pdf", width=12, height=8)
grid.draw(g1)
dev.off() 

pdf("bb_TRM_active_20TFs.pdf", width=12, height=5)
grid.draw(g2)
dev.off() 

pdf("bb_TEX_and_TRM_30TFs.pdf",width=12, height=7)
grid.draw(g3)
dev.off() 
```



## session info
```{r }
sessionInfo()
```

