---
title: "get_repressor_of_repressor"
output: html_document
date: "2025-02-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This script is to find the repressor of repressors in TexTerm. Here's how to do that:
1. find the genes which are highly expressed in TRM and lowly expressed in TexTerm, which are referred to as TRM functional genes
2. find the repressor of the functional genes in TexTerm. The TF scope is within the perturb-seq targets in chronic infection

## import data
```{r import data}
rna_seq <- read.csv('rna_all.csv', row.names = 1)
group_sorted <- read.csv('../Fig1/group_file_for_heatmap.csv',header = T, row.names = 1)
```

## identify TRM functional genes
```{r identify TRM functional genes}
# manually select background samples
group2 <- group_sorted[grepl("TRM|TexTerm",group_sorted$Group),]
x <- "TRM"
pr <- rna_seq[,rownames(group2)]
n <- which(group2[["Group"]]==x)
```


```{r identify TRM functional genes}
fl.sources <- list.files("../../scripts/utils/", full.names = T)
tmp <- sapply(fl.sources,source)
x <- unpairedtTest(pr,n,p=0.05,d=0.5)
tfs <- rownames(x)
```


```{r identify TRM functional genes}
writeLines(tfs, 'TRM_functional_genes_p0.05_d0.5.txt')
```

## identify repressor of the TRM functional genes
Expression of more than half of the TRM functional genes are reduced in knock-out TFs are identified as repressor. We'll use the perturb-seq seurat object in chronic infection as the main data
```{r identify rep}

```



