---
title: "comparison between two clustering methods"
author: "Cong"
date: "2025-12-16"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# import packages and functions

```{r import packages and functions}
suppressMessages(library(mclust))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(ggalluvial))
suppressMessages(library(tidyr))
```

# data loading

-   clustering 1: main manuscript clustering is based on correlation of correlation, "neighbor of neighbors"
-   clustering 2: more standard clustering based on correlation

```{r}
set.seed(123)
sample <- 'TRM' ### 'TRM' or 'TexTerm'
tag <- '_subset_TFs_v2'
df_old <- read.csv(paste0("clustering_membership_copula_",sample,"_shrinkage_res0.9_s0.15",tag,".csv"))
df_new <- read.csv(paste0("clustering_membership_copula_",sample,"_shrinkage_res1.1_s0.15",tag,".csv"))

```


# preprocessing & filtering

We merged two clustering into a single dataframe and removed the standalone clusters.

```{r}
# Merge the two results by TF name
# We suffix columns with _old and _new
df_compare <- inner_join(df_old, df_new, by = c("name", "group"), suffix = c("_old", "_new"))

# Function to identify clusters of size 1 (Standalone)
get_singleton_clusters <- function(cluster_vec) {
  counts <- table(cluster_vec)
  names(counts)[counts == 1]
}

# Identify singleton clusters in both methods
singletons_old <- get_singleton_clusters(df_compare$cluster_old)
singletons_new <- get_singleton_clusters(df_compare$cluster_new)

# Filter: Exclude TFs that belong to a singleton cluster in EITHER method
df_clean <- df_compare %>%
  filter(!cluster_old %in% singletons_old,
         !cluster_new %in% singletons_new)

cat("TFs remaining after removing standalone clusters:", nrow(df_clean), "\n")
```

We want to align two clusters first.
```{r}

# ============================================================
# FUNCTION: Align Clusters
# ============================================================
align_clusters <- function(data, old_col, new_col) {
  
  # 1. Create a contingency table (Confusion Matrix)
  # Rows = Old, Cols = New
  conf_matrix <- table(data[[old_col]], data[[new_col]])
  
  # 2. Convert to a formatted dataframe for sorting
  matches <- as.data.frame(conf_matrix)
  colnames(matches) <- c("Old", "New", "Count")
  
  # Sort by Count descending (Biggest overlaps first)
  matches <- matches %>% arrange(desc(Count))
  
  # 3. Create a mapping vector
  # We need to ensure we don't map the same cluster twice
  mapping <- list()
  used_old <- c()
  used_new <- c()
  
  for(i in 1:nrow(matches)) {
    row <- matches[i, ]
    o <- as.character(row$Old)
    n <- as.character(row$New)
    
    # If neither cluster has been assigned yet, map them!
    if(!(o %in% used_old) && !(n %in% used_new)) {
      mapping[[n]] <- o
      used_old <- c(used_old, o)
      used_new <- c(used_new, n)
    }
  }
  
  # 4. Handle any leftover clusters (if counts differ)
  # Assign remaining New clusters to their original names or unassigned numbers
  all_new <- unique(as.character(data[[new_col]]))
  remaining_new <- setdiff(all_new, names(mapping))
  
  for(r in remaining_new) {
    mapping[[r]] <- paste0(r, "_unmatched") 
  }
  
  # 5. Apply the mapping to the dataframe
  # Create a new column called 'cluster_new_aligned'
  data$cluster_new_aligned <- sapply(as.character(data[[new_col]]), function(x) mapping[[x]])
  
  # Convert to factor to ensure order matches Old clusters
  # This makes the diagonal plotting work automatically
  data$cluster_new_aligned <- factor(data$cluster_new_aligned, levels = levels(factor(data[[old_col]])))
  
  return(data)
}

# ============================================================
# EXECUTION
# ============================================================

# Run the alignment
df_clean <- align_clusters(df_clean, "cluster_old", "cluster_new")

# Check the results
print("Mapping Table (Visual Check):")
table(df_clean$cluster_old, df_clean$cluster_new_aligned)

# Now use 'cluster_new_aligned' for your plots!
# When you plot this table, the heavy numbers will be on the diagonal.
```


We manually checked some important genes. All remained stable clustering.
```{r}
tmp <- c("Fosb","Snai1","Klf6","Irf8","Hic1","Prdm1","Gfi1","Fli1")
knitr::kable(df_clean[df_clean$name %in% tmp,], caption = 'clustering comparison on key TFs') |> kableExtra::kable_styling(latex_options = 'scale_down')

```

# global comparison (level 1)

We compared the global clustering assignment regardless of TF group
```{r}
# A. Quantitative Metric: Adjusted Rand Index (ARI)
# ARI = 1.0 means perfect match. ARI ~ 0 means random chance.
global_ari <- adjustedRandIndex(df_clean$cluster_old, df_clean$cluster_new_aligned)
cat("Global Adjusted Rand Index (ARI):", round(global_ari, 3), "\n")

# B. Visualization: Confusion Matrix Heatmap
# This shows: "Did TFs in Old Cluster 1 move to New Cluster 3?"
confusion_mat <- df_clean %>%
  count(cluster_old, cluster_new_aligned)

p1 <- ggplot(confusion_mat, aes(x = factor(cluster_new_aligned), y = factor(cluster_old), fill = n)) +
  geom_tile() +
  geom_text(aes(label = n), color = "white") +
  scale_fill_viridis_c() +
  labs(title = paste0("Global Cluster Comparison (ARI = ", round(global_ari, 3), ")"),
       x = "New Method Clusters",
       y = "Old Method Clusters",
       fill = "# TFs") +
  theme_minimal()

print(p1)

```

# group-wise comparison (level 2)

We compared the clustering within each TF group
```{r}
# A. Calculate ARI per TF Group
group_metrics <- df_clean %>%
  group_by(group) %>%
  summarise(
    n_tfs = n(),
    group_ari = adjustedRandIndex(cluster_old, cluster_new_aligned)
  )

print(group_metrics)
```

Tex_important and Tex_specific groups remained highly stable.
```{r}
# B. Visualization: Bar Chart of Stability per Group
p2 <- ggplot(group_metrics, aes(x = reorder(group, group_ari), y = group_ari, fill = group)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  ylim(0, 1) +
  labs(title = "Cluster Stability by TF Group",
       subtitle = "Higher ARI = Clusters remained more consistent within this group",
       y = "Adjusted Rand Index (Agreement)",
       x = "TF Group") +
  theme_classic()

print(p2)  
  
```

If you see straight ribbons flowing from Left to Right, the clusters are stable. If the ribbons criss-cross violently, the TFs were reassigned.

```{r}
# C. Visualization: Detailed Flow (Alluvial Plot) per Group
p3 <- ggplot(df_clean,
       aes(axis1 = factor(cluster_old), axis2 = factor(cluster_new_aligned))) +
  geom_alluvium(aes(fill = group), width = 1/12) +
  geom_stratum(width = 1/12, fill = "grey80", color = "grey") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Old Clusters", "New Clusters"), expand = c(.05, .05)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "TF Cluster Flow: Old vs New",
       subtitle = "Colored by Biological Group") +
  theme_minimal() +
  facet_wrap(~group, scales = "free") # Facet by group to see individual alignments

print(p3)
```


# session info

```{r }
sessionInfo()
```
