---
title: "context-dependent TF cooperation network construction"
author: "Cong"
date: "2025-12-16"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# import packages and functions

```{r import packages and functions}
suppressMessages(library(igraph))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(RColorBrewer))
suppressMessages(library(huge))
suppressMessages(library(xlsx))

set.seed(42)
fl.sources <- list.files("../../scripts/utils/", full.names = T)
tmp <- sapply(fl.sources,source)
```

# prepare TF-gene edge weight input

Since each context (TRM or TexTerm) has several networks in different replicate samples, we first calculated the mean edge weight of TF-regulatee pair across samples. We kept regulatees with moderate variation across TFs (sd\>1).

-   input: `edge_combined.csv` in each `network/` folder
-   optional argument: pre-defined TF list, here, we only focus on TFs important in each context
-   output file: `mean_edge_weight.csv`: rows are regulatees (regulated downstream genes) and columns are TFs
We'll skip this part for demo purpose. The output file is provided.

```{r}
# sample <- 'TRM' # 'TRM' or 'TexTerm'
# tag <- '_subset_TFs_v2'
# 
# L <- readLines(paste0(sample,'_samples.txt'))
# TFs <- readLines(paste0(sample,'_TFs_for_network_v3.txt')) 
# TFs <- lapply(TFs, toupper)
#  
# 
# edge_weight <- do.call("rbind", lapply(L, function(x) read.table(paste0(x,'/edges_combined.csv'), sep = ",", header = T) %>% select(c(1,2,3)) %>% setNames(c("TF","regulatee","weight")))) %>% dplyr::filter(TF %in% TFs) %>% 
#                group_by(TF, regulatee) %>% dplyr::summarize(weight = mean(weight, na.rm=TRUE)) %>% 
#                reshape2::dcast(regulatee ~ TF, value.var = "weight", fill=0) %>% 
#                tibble::column_to_rownames(var = "regulatee") 
#                                        
# df <- edge_weight %>% rowwise() %>% mutate(sd=sd(c_across(everything()))) %>% ungroup() %>% 
#                 filter(sd>1) %>% select(!matches("MOUSE$|[0-9\\.]{6}$")) %>% select(-sd) %>%
#                 rename_with(firstup)
# write.csv(df,file = paste0("mean_",sample,"_edge_weight",tag,".csv")) # write to file
```

```{r}
sample <- 'TRM' ### 'TRM' or 'TexTerm'
tag <- '_subset_TFs_v2'
df <- read.csv(paste0("mean_",sample,"_edge_weight",tag,".csv"), row.names = 1)
colnames(df) <- firstup(colnames(df))
knitr::kable(df[10000:10006,1:6], caption = 'TF-gene edge weight') |> kableExtra::kable_styling(latex_options = 'scale_down')
print(dim(df))
```

# copula pipeline to infer graph

We used R package `huge` to construct Gaussian graphical models (GGM) with copula modification. A little bit background to help with selction of hyper-parameters.

GGM assumes that the observations have a multivariate Gaussian distribution with mean $µ$, and covariance matrix $Σ$. The conditional independence can be implied by the inverse covariance (concentration) matrix $Ω = Σ^{−1}$. If $Ω_{jk} = 0$, then the i-th variable and j-th variables are conditional independent given all other variables.

This important property serves as the foundation for GGM to infer direct interactions from high-dimensional data. Unlike relevance networks or correlation networks, in which edges are determined based on marginal correlations, GGM provides a stronger criterion of dependency, and thus further reduces the false positive rate.

However, a great limitation of classic learning methods for GGM is the lack of sparsity in the resulting graph. A dense graph not only complicates downstream analysis but also raises the issue of overfitting the data. Thus it makes sense to impose an $ℓ$ penalty for the estimation of $Ω$, to increase its sparsity, and the sparse pattern of $Ω$ is essentially the same as the adjacency matrix of the underlying undirected graph.

Additionally, [Liu et al.](https://projecteuclid.org/journals/annals-of-statistics/volume-40/issue-4/High-dimensional-semiparametric-Gaussian-copula-graphical-models/10.1214/12-AOS1037.full) developed a data transformation method called Copula that can be used with the graphical Lasso algorithm to relax the normality assumption of GGM. The intuition of Copula is to infer the the relationship of the variables without being influenced by their individual distributions

For more details about package `huge`, check the [vignette](https://cran.r-project.org/web/packages/huge/vignettes/vignette.pdf). For instructions on selecting the hyper-parameters like `npn.func`, check the [manual](https://cran.r-project.org/web/packages/huge/huge.pdf)

```{r}
npn.func <- 'shrinkage'
df.npn <- huge.npn(df, npn.func = npn.func) # default                      
out.npn <- huge(df.npn, method = "glasso", nlambda = 30, lambda.min.ratio = 0.01) # fit glasso model to data
saveRDS(out.npn,paste0("result_copula_",sample,'_',npn.func,tag,".rds"))
```

The program automatically sets up a sequence of 30 regularization parameters and estimates the corresponding graph path.

```{r, fig.width=8, fig.height=5}
# default visualization provided by huge package
plot(out.npn)
```

## get precision matrix

From the above plot, we can see the local minimum of sparsity is 0.15. We can infer the precision matrix when sparsity=0.15

```{r}
sparsity <- 0.15
idx = which(out.npn$sparsity>=sparsity & out.npn$sparsity<=sparsity+0.01)[1]
print(paste0("lambda: ", out.npn$lambda[idx]))
print(paste0("sparsity: ", out.npn$sparsity[idx]))
```


```{r}
# 1. Get the Precision Matrix (Inverse Covariance)
theta <- as.matrix(out.npn$icov[[idx]])

# 2. Convert Precision Matrix to Partial Correlation Matrix
# Formula: rho_ij = - theta_ij / sqrt(theta_ii * theta_jj)
diag_theta <- diag(theta)
m <- -theta / sqrt(outer(diag_theta, diag_theta, "*"))
diag(m) <- 1 # The diagonal of a correlation matrix is always 1

# 3. Apply names
rownames(m) <- colnames(df)
colnames(m) <- colnames(df)

knitr::kable(head(m[,1:6]), caption = 'sparse TF-TF correlation matrix') |> kableExtra::kable_styling(latex_options = 'scale_down')

## write to file
write.csv(m, paste0("partial_correlation_copula_",sample,"_",npn.func,"_sparsity_",sparsity,tag,".csv"))
```

# create igraph plot

## create dataframe for igraph

```{r}
df <- data.frame(row=rownames(m)[row(m)[upper.tri(m)]], 
                 col=colnames(m)[col(m)[upper.tri(m)]], 
                 corr=m[upper.tri(m)])
### remove zero entries
df <- df[df$corr!=0,]
knitr::kable(head(df), caption = 'long form of TF correlation') |> kableExtra::kable_styling(latex_options = 'scale_down')

print(dim(df))
print(paste('percent of correlation score >','0',':',sum(df$corr > 0) *100 / nrow(df),"%"))
print(paste('percent of correlation score <','0',':',sum(df$corr < 0) *100 / nrow(df),"%"))

df$corr = abs(df$corr)
write.csv(df, paste0("correlation_intensity_table_copula_",sample,"_",npn.func,"_sparsity_",sparsity,tag,".csv"))
```

## import node meta

Node annotation came from the results of Fig. S3a.

-   `TRM_Tex`: multi-tasker in both TRM and TexTerm
-   `Tex_specific`/`TRM_specific`: single-tasker in TexTerm/TRM
-   `Tex_important`/`TRM_important`: important in TexTerm/TRM
-   `Housekeeping`: 54 universal TFs with small variation across cell types

```{r}
nodes <- read.xlsx('TRM_Tex_TF_list_20240121.xlsx', sheetName = sample) %>% tidyr::drop_na(TF) %>% dplyr::mutate(group=ifelse(is.na(Specificity), Important, Specificity)) %>% select(TF, group)
names(nodes) <- c('name','group')
knitr::kable(head(nodes), caption = 'node meta') |> kableExtra::kable_styling(latex_options = 'scale_down')
print(dim(nodes))   
  
```

## graph generation

```{r}
g = graph_from_data_frame(df, directed=FALSE, vertices = nodes)
print(g, e=TRUE, v=TRUE)
```

## Leiden clustering

Leiden clustering generates communities. Adjust the `res` to get a reasonable number of groups and high modularity score. High `res` gives more groups. Through experimenting, res=0.9 suits the best.

```{r}
set.seed(42)
res <- 1.1
clustering = cluster_leiden(g, objective_function="modularity",resolution_parameter=res)
print(clustering)
print(sizes(clustering))
print(paste0('modularity:',modularity(g, membership(clustering))))
saveRDS(clustering,paste0('clustering_copula_',sample,'_',npn.func,'_res',res,'_s',sparsity,tag,'.rds'))
```

```{r}
## write clustering result
clust <- data.frame(name = names(membership(clustering)),cluster = as.numeric(membership(clustering))) %>%
        dplyr::inner_join(nodes, by = "name") %>%
        dplyr::arrange(group, cluster)
knitr::kable(head(clust), caption = 'community clustering') |> kableExtra::kable_styling(latex_options = 'scale_down')
write.csv(clust,paste0('clustering_membership_copula_',sample,'_',npn.func,'_res',res,'_s',sparsity,tag,'.csv'), row.names=F)


### save graph
saveRDS(g, paste0("graph_",npn.func,'_res',res,"_s",sparsity,"_",sample,"_",tag,".rds"))

```

```{r }
sessionInfo()
```
