---
title: "visualization of context-dependent TF-TF cooperation network"
author: "Cong"
date: "2024-07-03"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# import packages and functions
```{r import packages and functions}
suppressMessages(library(igraph))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(RColorBrewer))

```

# load two networks first
```{r}
g1 <- readRDS('graph_shrinkage_res0.9_s0.15_TexTerm_subset_TFs_v2.rds')
g2 <- readRDS('graph_shrinkage_res0.9_s0.15_TRM_subset_TFs_v2.rds')
gv1 <- igraph::as_data_frame(g1, "vertices")
gv2 <- igraph::as_data_frame(g2, "vertices")

common_tfs <- intersect(gv1$name, gv2$name)
tex_tfs <- setdiff(gv1$name, gv2$name)
trm_tfs <- setdiff(gv2$name, gv1$name)



```

# combine two networks
first we prepared the edge info
```{r}
# extract edge info: take mean if edge exists in both networks
ge1 <- igraph::as_data_frame(g1) %>% select(from, to, corr) %>% transform(from=pmin(from, to), to=pmax(from,to))
ge2 <- igraph::as_data_frame(g2) %>% select(from, to, corr) %>% transform(from=pmin(from, to), to=pmax(from,to))
ge <- ge1 %>% full_join(ge2, by = c('from','to')) %>% 
    mutate(corr=ifelse(is.na(corr.x), corr.y, ifelse(is.na(corr.y), corr.x, (corr.x+corr.y)/2)), 
           edge.color=ifelse(is.na(corr.x), '#009051', ifelse(is.na(corr.y), '#8AAA75', '#808080')),
           type=ifelse(is.na(corr.x), 'trm', ifelse(is.na(corr.y), 'tex', 'common'))) |> 
        select(from,to,corr,edge.color,type) |> na.omit()
knitr::kable(head(ge), caption = 'edge meta') |> kableExtra::kable_styling(latex_options = 'scale_down') 
```

Then we prepared the node info. Most of the TFs are in the same community in both contexts. Some TFs are different so they will be marked as "multi"
```{r get node info}
gv1 <- igraph::as_data_frame(g1, "vertices") %>% select(name, cluster)
gv2 <- igraph::as_data_frame(g2, "vertices") %>% mutate(cluster2=ifelse(cluster==1, 3, ifelse(cluster==3, 1, cluster)))%>% select(name, cluster2)
attrs2 <- gv1 %>% full_join(gv2, by=c('name')) %>% 
        mutate(group=ifelse(name %in% common_tfs, 'Common', ifelse(name %in% tex_tfs, 'Tex', 'Trm')), 
               color=ifelse(name %in% common_tfs, '#C0C0C0', ifelse(name %in% tex_tfs, '#BB8666', '#98D7AD')),
               cluster_c=ifelse(is.na(cluster),cluster2,ifelse(is.na(cluster2), cluster,ifelse(cluster==cluster2, cluster, 'mixed'))))
knitr::kable(head(attrs2), caption = 'node meta') |> kableExtra::kable_styling(latex_options = 'scale_down') 

```


```{r create combined network}
# create combined network from the edge meta
g <- graph_from_data_frame(ge, directed = FALSE, vertices = attrs2)
print(g, e=TRUE, v=TRUE)
```

# network visualization
## fix the network layout
```{r}
G_Grouped = g
E(G_Grouped)$weight = 1

## Add edges with high weight between all nodes in the same group
for(i in unique(V(g)$cluster_c)) {
    GroupV = which(V(g)$cluster_c == i)
    G_Grouped = add_edges(G_Grouped, combn(GroupV, 2), attr=list(weight=1.8))
} 

## Now create a layout based on G_Grouped
set.seed(567)
LO = layout_with_fr(G_Grouped)
```

## define function
```{r}
# function to generate network
get_figure <- function(x,ge,cut_off=0.8,label=F){
    g <- graph_from_data_frame(ge, directed = FALSE, vertices = attrs2)
    G_Grouped = g
    E(G_Grouped)$weight = 1
    
    ## Add edges with high weight between all nodes in the same group
    for(i in unique(V(g)$cluster_c)) {
        GroupV = which(V(g)$cluster_c == i)
        G_Grouped = add_edges(G_Grouped, combn(GroupV, 2), attr=list(weight=1.8))
    } 
    
    ## Now create a layout based on G_Grouped
    set.seed(567)
    LO = layout_with_fr(G_Grouped)
    
    # get degree
    ge2 <- ge |> dplyr::filter(type==x)
    g2 <- graph_from_data_frame(ge2, directed = FALSE, vertices = attrs2) # to get customized node size in each context
    # generate label for all nodes
    V(g)$frame.color <- 'gray'
    V(g)$label <- NA
    if (label){V(g)$label <- V(g)$name}
    
    V(g)$label.color <- 'black'
    V(g)$size <- 0.2*degree(g2) # degree in each context
    # V(g)$label.cex <- 0.75 # part label
    V(g)$label.cex <- 0.5 # all label
    
    # set distance of label
    V(g)$label.dist <- 0.4
    
    # set edge color
    # E(g)$color <- "gray80"
    E(g)$color <- ifelse(E(g)$type=='common', '#C0C0C0', ifelse(E(g)$type=='trm', '#009051', '#945200'))
    E(g)$size <- 0.1
    E(g)$width <- 2*E(g)$corr
    E(g)$lty <- ifelse((E(g)$corr>=unname(quantile(E(g)$corr, cut_off))) & (E(g)$type==x), 1, 0)

    plot(g, layout=LO, main = paste0(x, ' network'))
}
```

## generate figures
```{r, fig.width=8, fig.height=8}
tmps <- lapply(c('common','trm','tex'), function(x) get_figure(x=x, ge=ge, cut_off=0.8,label=T))

```

# individual TF's neighbors in different contexts
```{r}
# define function
get_graph <- function(id){
    ge1 <- igraph::as_data_frame(g1) %>% select(from, to, corr) %>% transform(from=pmin(from, to), to=pmax(from,to)) %>% mutate(edge.color='#7F6000')
    ge2 <- igraph::as_data_frame(g2) %>% select(from, to, corr) %>% transform(from=pmin(from, to), to=pmax(from,to)) %>% mutate(edge.color='#009051')
    ge <- rbind(ge1,ge2)
    
    ge3 <- ge %>% filter(from == id | to == id)
    nodes <- union(ge3$from, ge3$to)
    attrs3 = data.frame('name'=nodes) %>% mutate(color=ifelse(name %in% common_tfs, '#C0C0C0', ifelse(name %in% tex_tfs, '#7F6000', '#009051')))

    g <- graph_from_data_frame(ge3, directed = FALSE, vertices = attrs3)
    V(g)$label.color <- 'black'
    # V(g)$size <- 10*degree(g)
    V(g)$size <- ifelse(V(g)$name == id, 12, 7)
    # V(g)$label.cex <- 0.75 # part label
    V(g)$label.cex <- 1 # all label

    # set distance of label
    V(g)$label.dist <- 1

    # set edge color
    # E(g)$color <- "gray80"
    E(g)$color <- ge3$edge.color
    E(g)$size <- 1
    E(g)$width <- 3*E(g)$corr
    G_Grouped = g
    E(G_Grouped)$weight = 1

    ## make edges of the same type close to each other in the layout
    for(i in unique(V(g)$color)) {
        GroupV = which(V(g)$color == i)
        G_Grouped = add_edges(G_Grouped, combn(GroupV, 2), attr=list(weight=0.1))
    } 

    ## Now create a layout based on G_Grouped
    set.seed(567)
    LO = layout_with_fr(G_Grouped)    
    
    set.seed(1)
    plot(g, layout=LO, main = paste0(id, ' network'))

}

```

Now we can visualize some TFs' neighbors in both TRM and TEXterm. Hic1 is a common TF in both contexts. Zfp324 is TEXterm-specific TF. Fosb is TRM-specific TF. 
```{r, fig.width=6, fig.height=6}
# visualize some example TFs
tmp <- lapply(c('Hic1','Fosb','Zfp324'), get_graph)


```


# session info
```{r }
sessionInfo()
```

